Quantum Interval-Valued Probability: Contextuality and the Born Rule

### 2017-08-22 prop_letter.tex
pass over sec 4

Our generalization of quantum probability measures to QIVPMs allows us
to strengthen one of the fundamental theorems in of quantum
information: the Kochen-Specker
theorem~\cite{PeresRon1988,peres1995quantum}. Our stronger version of
the theorem will suggest a resolution to the debate initiated by Meyer
and Mermin on the relevance of the Kochen-Specker to experimental, and
hence finite-precision, quantum
measurements~\citep{PhysRevLett.83.3751,Mermin1999,Kent1999,HavlicekKrennSummhammerSvozil2001,SimonBruknerZeilinger2001,Cabello2002,Larsson2002,Appleby2002,Peres2003,BarrettKent2004,Appleby_2005,Spekkens2005,GuehneKleinmannCabelloEtAl2010,MazurekPuseyKunjwalEtAl2016}.
Specifically, the original Kochen-Specker theorem is formulated under
the assumption that all observables of a quantum mechanical system
have \emph{definite values at all times}~\cite{Held2016}. Our
interval-valued probability framework will allow us to state, and prove, a
stronger version of the theorem that holds even if the observables
have values that are only definite up to some specific parameter that
quantifies experimental uncertainty. Put differently, our stronger
statement and theorem quantifies Mermin's intuition:
\begin{quote}
  \ldots although the outcomes deduced from such imperfect
  measurements will occasionally differ dramatically from those
  allowed in the ideal case, if the misalignment is very slight, the
  statistical distribution of outcomes will differ only slightly from
  the ideal case~\citep{Mermin1999}.
\end{quote}

@@

The first step in our formalization
is to introduce a family of QIVPMs parameterized by an uncertainty
$\delta$ which we call $\delta$-deterministic QIVPMs.

@@

If a QIVPM is 0-deterministic, then for a family of commuting
observables~$\left\{ \mathbf{O}_{j}\right\} _{j=1}^{M}$, we have:
\begin{equation}
\prod_{j=1}^{M}\expval{\mathbf{O}_{j}}_{\bar{\mu}}=\expval{\prod_{j=1}^{M}\mathbf{O}_{j}}_{\bar{\mu}}\label{eq:Kochen-Specker-Rules}
\end{equation}
Intuitively the expectation value of an observable lies between two
possible outcomes, which themselves lie between the minimum and
maximum bounds of the given probability intervals.  If the QIVPM is
0-deterministic, this interval collapses to a singleton
$\left\{ \lambda\right\} $, where $\lambda$ is an eigenvalue
of~$\mathbf{O}$.

@@

\begin{thm}
[0-Deterministic Variant of Kochen-Specker] \label{thm:Kochen-Specker} 
Given a Hilbert space $\Hilb$ of dimension~$d\ge3$, there is no
0-deterministic
measure~$\mu^{\mathrm{D}}:\events\rightarrow\left\{ 0,1\right\}$.
\end{thm}

To prove this result, we proceed by assuming a 0-deterministic QIVPM
$\bar{\mu}$ and derive a contradiction using the same nine observables
$\mathbf{O}_{ij}$ used in the ``Mermin-Peres'' proof of the
Kochen-Specker theorem~\cite{Mermin1990Simple,peres1995quantum}:
{\renewcommand{\arraystretch}{2}%
\begin{center}
\begin{tabular}{|@{\quad}c@{\quad}|@{\quad}c@{\quad}|@{\quad}c@{\quad}|}
\hline 
$\mathbb{1}\otimes\sigma_{z}$  & $\sigma_{z}\otimes\mathbb{1}$  & $\sigma_{z}\otimes\sigma_{z}$ \tabularnewline
\hline 
$\sigma_{x}\otimes\mathbb{1}$  & $\mathbb{1}\otimes\sigma_{x}$  & $\sigma_{x}\otimes\sigma_{x}$ \tabularnewline
\hline 
$\sigma_{x}\otimes\sigma_{z}$  & $\sigma_{z}\otimes\sigma_{x}$  & $\sigma_{y}\otimes\sigma_{y}$ \tabularnewline
\hline 
\end{tabular}\,,
\par\end{center}
}
\noindent where $\left\{\mathbb{1},\sigma_{x},\sigma_{y},\sigma_{z}\right\}$ are
Pauli matrices acting on a pair of spin-$\frac{1}{2}$ states. We now define the
expectation value over the QIVPM for our Mermin-Peres square.

Since the expectation value~$\expval{\mathbf{O}_{ij}}_{\bar{\mu}}$ satisfies
these two properties, the observables in the Mermin-Peres square must
have the expectation value~$\left\{ 1\right\} $ or
$\left\{ -1\right\} $.  In each row and column, the operators commute,
and each operator is the product of two others, except in the third
column, where
$\left(\sigma_{z}\otimes\sigma_{z}\right)\left(\sigma_{x}\otimes\sigma_{x}\right)=-\sigma_{y}\otimes\sigma_{y}$.
Therefore, it is clearly impossible to find a QIVPM~$\bar{\mu}$ such
that $\langle \mathbf{O}_{ij}\rangle_{\bar \mu}$ satisfies
Eq.~(\ref{eq:Kochen-Specker-Rules}).

@@

%\begin{proof} 
Suppose there is a $\delta$-deterministic
  QIVPM~$\bar{\mu}:\events\rightarrow\mathscr{I}$, we claim that the
  following function~$\bar{\mu}^{\textrm{D}}:\events\rightarrow\left\{
    \imposs,\necess\right\} $ is a cryptodeterministic QIVPM:
\begin{equation}
\bar{\mu}^{\textrm{D}}\left(P\right)=\begin{cases}
\imposs\,, & \textrm{ if }\bar{\mu}\left(P\right)\subseteq\left[0,\delta\right]\:;\\
\necess\,, & \textrm{ if }\bar{\mu}\left(P\right)\subseteq\left[1-\delta,1\right]\:.
\end{cases}
\end{equation}
However, since we know that there is no cryptodeterministic QIVPM,
$\bar{\mu}^{\textrm{D}}$ cannot exist and hence $\bar{\mu}$ cannot
exist.

It remains to verify that $\bar{\mu}^{\textrm{D}}$ is a cryptodeterministic
QIVPM by checking the properties of
\begin{subequations}
\begin{eqnarray}
 &  & \bar{\mu}^{\textrm{D}}\left(P_{0}\right)=\bar{\mu}^{\textrm{D}}\left(P_{1}\right)=\imposs\\
 & \Rightarrow & \bar{\mu}^{\textrm{D}}\left(P_{0}+P_{1}\right)=\imposs
\end{eqnarray}
\end{subequations}for orthogonal operators~$P_{0}$ and $P_{1}$.

When $\bar{\mu}^{\textrm{D}}\left(P_{0}\right)=\bar{\mu}^{\textrm{D}}\left(P_{1}\right)=\imposs$,
we have both $\bar{\mu}\left(P_{0}\right)$ and $\bar{\mu}\left(P_{1}\right)\subseteq\left[0,\delta\right]$
which implies $\bar{\mu}\left(P_{0}+P_{1}\right)\subseteq\left[0,2\delta\right]$.
Since $\delta<\frac{1}{3}\Rightarrow2\delta<1-\delta$, $\bar{\mu}\left(P_{0}+P_{1}\right)$
cannot be a subset of $\left[1-\delta,1\right]$, but $\bar{\mu}\left(P_{0}+P_{1}\right)$
is a subset of either $\left[0,\delta\right]$ or $\left[1-\delta,1\right]$.
Hence, $\bar{\mu}\left(P_{0}+P_{1}\right)$ must be a subset of $\left[0,\delta\right]$,
which implies $\bar{\mu}^{\textrm{D}}\left(P_{0}+P_{1}\right)=\imposs$. Thus the
theorem holds when $\delta<\frac{1}{3}$.
%\end{proof}

%%%%%
\subsection{HV}

A $\frac{1}{3}$-deterministic QIVPM can be given by
$\bar{\mu}\left(P\right)=\left[n/3,n/3\right]$, where $P$ projects
onto an $n$-dimensional subspace of a three-dimensional Hilbert space.

>>>>

Our generalization of quantum probability measures to QIVPMs allows us
to strengthen one of the fundamental theorems of quantum information:
the Kochen-Specker theorem~\cite{PeresRon1988,peres1995quantum}. Our
stronger version of the theorem will suggest a resolution to the
debate initiated by Meyer and Mermin on the relevance of the
Kochen-Specker to experimental, and hence finite-precision, quantum
measurements~\citep{PhysRevLett.83.3751,Mermin1999,Kent1999,HavlicekKrennSummhammerSvozil2001,SimonBruknerZeilinger2001,Cabello2002,Larsson2002,Appleby2002,Peres2003,BarrettKent2004,Appleby_2005,Spekkens2005,GuehneKleinmannCabelloEtAl2010,MazurekPuseyKunjwalEtAl2016}.
Specifically, the original Kochen-Specker theorem is formulated under
the assumption that all observables of a quantum mechanical system
have \emph{definite values at all times}~\cite{Held2016}. Our
interval-valued probability framework will allow us to state, and
prove, a stronger version of the theorem that holds even if the
observables have values that are only definite up to some specific
parameter that quantifies experimental uncertainty. Put differently,
our stronger statement and theorem quantifies Mermin's intuition:
\begin{quote}
  \ldots although the outcomes deduced from such imperfect
  measurements will occasionally differ dramatically from those
  allowed in the ideal case, if the misalignment is very slight, the
  statistical distribution of outcomes will differ only slightly from
  the ideal case~\citep{Mermin1999}.
\end{quote}

@@

The first step in our formalization is to introduce a family of QIVPMs
parameterized by an uncertainty~$\delta$ which we call
$\delta$-deterministic QIVPMs.

@@

If a QIVPM is 0-deterministic, then for a family of commuting
observables~$\left\{ \mathbf{O}_{j}\right\} _{j=1}^{M}$, we can prove
that the product of expectation values of the individual observables
is the expectation value of the product of the family:
\begin{equation}
\prod_{j=1}^{M}\expval{\mathbf{O}_{j}}_{\bar{\mu}}=\expval{\prod_{j=1}^{M}\mathbf{O}_{j}}_{\bar{\mu}}\label{eq:Kochen-Specker-Rules}
\end{equation}
\amr{Talk with Yu-Tsung about this} Intuitively the expectation value
of each observable lies between two possible outcomes, which
themselves lie between the minimum and maximum bounds of the given
probability intervals.  If the QIVPM is 0-deterministic, this interval
collapses to a singleton $\left\{ \lambda\right\} $, where $\lambda$
is an eigenvalue of~$\mathbf{O}$.

@@

\begin{thm} [0-Deterministic Variant of
  Kochen-Specker] \label{thm:Kochen-Specker} Given a Hilbert space
  $\Hilb$ of dimension~$d\ge3$, there is no 0-deterministic
  measure~$\bar{\mu}$ mapping every event to either $[0,0]$ or $[1,1]$.
\end{thm}

To explain why this result is equivalent to the original
Kochen-Specker theorem and to prove it at the same time, we proceed by
assuming a 0-deterministic QIVPM $\bar{\mu}$ and derive a
contradiction to the original Kochen-Specker theorem using the same
nine observables $\mathbf{O}_{ij}$ used in the ``Mermin-Peres'' proof
of the Kochen-Specker
theorem~\cite{Mermin1990Simple,peres1995quantum}:
{\renewcommand{\arraystretch}{2}%
\begin{center}
\begin{tabular}{|@{\quad}c@{\quad}|@{\quad}c@{\quad}|@{\quad}c@{\quad}|}
\hline 
$\mathbb{1}\otimes\sigma_{z}$  & $\sigma_{z}\otimes\mathbb{1}$  & $\sigma_{z}\otimes\sigma_{z}$ \tabularnewline
\hline 
$\sigma_{x}\otimes\mathbb{1}$  & $\mathbb{1}\otimes\sigma_{x}$  & $\sigma_{x}\otimes\sigma_{x}$ \tabularnewline
\hline 
$\sigma_{x}\otimes\sigma_{z}$  & $\sigma_{z}\otimes\sigma_{x}$  & $\sigma_{y}\otimes\sigma_{y}$ \tabularnewline
\hline 
\end{tabular}\,,
\par\end{center}
}
\noindent where
$\left\{\mathbb{1},\sigma_{x},\sigma_{y},\sigma_{z}\right\}$ are Pauli
matrices acting on a pair of spin-$\frac{1}{2}$ states.  The
observables in the Mermin-Peres square must have expectation
values~$\left\{ 1\right\} $ or $\left\{ -1\right\} $.  In each row and
column, the operators commute, and each operator is the product of two
others, except in the third column, where
$\left(\sigma_{z}\otimes\sigma_{z}\right)\left(\sigma_{x}\otimes\sigma_{x}\right)=-\sigma_{y}\otimes\sigma_{y}$.
Therefore, it is clearly impossible to find a QIVPM~$\bar{\mu}$ such
that $\langle \mathbf{O}_{ij}\rangle_{\bar \mu}$ satisfies
Eq.~(\ref{eq:Kochen-Specker-Rules}).

We can in fact generalize the above theorem to state than for small
enough $\delta$, it is impossible to have just the two intervals
\imposs\ and \necess: every QIVPM must be parameterized by a
collection of intervals that includes at least one additional
``fuzzy'' interval and some events must be associated with such
uncertain intervals. The significance of this observation is that the
uncertainty in the intervals is, in contrast to the classical case,
\emph{a fundamental part of quantum theory}. There are in fact two
sources of uncertainty: one due to the finite measurement resources
and one due to complementarity, which requires every quantum framework
to be truly probabilistic.

@@

For the proof, suppose there is a $\delta$-deterministic
QIVPM~$\bar{\mu}:\events\rightarrow\mathscr{I}$, we claim that the
following
map~$\bar{\mu}^{\textrm{D}}:\events\rightarrow\left\{
  \imposs,\necess\right\} $ is a 0-deterministic QIVPM:
\begin{equation}
\bar{\mu}^{\textrm{D}}\left(P\right)=\begin{cases}
\imposs\,, & \textrm{ if }\bar{\mu}\left(P\right)\subseteq\left[0,\delta\right]\:;\\
\necess\,, & \textrm{ if }\bar{\mu}\left(P\right)\subseteq\left[1-\delta,1\right]\:.
\end{cases}
\end{equation}
However, since we know that 0-deterministic QIVPMs do not exist, the
map $\bar{\mu}^{\textrm{D}}$ cannot exist and hence $\bar{\mu}$ does
not exist.

It remains to verify that $\bar{\mu}^{\textrm{D}}$ is indeed a 0-deterministic
QIVPM by checking the following property:
\begin{subequations}
\begin{eqnarray}
 &  & \bar{\mu}^{\textrm{D}}\left(P_{0}\right)=\bar{\mu}^{\textrm{D}}\left(P_{1}\right)=\imposs\\
 & \Rightarrow & \bar{\mu}^{\textrm{D}}\left(P_{0}+P_{1}\right)=\imposs
\end{eqnarray}
\end{subequations}for orthogonal operators~$P_{0}$ and $P_{1}$.

When $\bar{\mu}^{\textrm{D}}\left(P_{0}\right)=\bar{\mu}^{\textrm{D}}\left(P_{1}\right)=\imposs$,
we have both $\bar{\mu}\left(P_{0}\right)$ and $\bar{\mu}\left(P_{1}\right)\subseteq\left[0,\delta\right]$
which implies $\bar{\mu}\left(P_{0}+P_{1}\right)\subseteq\left[0,2\delta\right]$.
Since $\delta<\frac{1}{3}\Rightarrow2\delta<1-\delta$, $\bar{\mu}\left(P_{0}+P_{1}\right)$
cannot be a subset of $\left[1-\delta,1\right]$, but $\bar{\mu}\left(P_{0}+P_{1}\right)$
is a subset of either $\left[0,\delta\right]$ or $\left[1-\delta,1\right]$.
Hence, $\bar{\mu}\left(P_{0}+P_{1}\right)$ must be a subset of $\left[0,\delta\right]$,
which implies $\bar{\mu}^{\textrm{D}}\left(P_{0}+P_{1}\right)=\imposs$. Thus the
theorem holds when $\delta<\frac{1}{3}$.

When $\delta \ge \frac{1}{3}$, the situation abruptly
changes. Assuming a Hilbert space $\Hilb$ of dimension $\ge 3$, then
for any projector $P$ that projects onto an $n$-dimensional subspace
of $\Hilb$, it is straightforward to check that
$\bar{\mu}\left(P\right)=\left[n/3,n/3\right]$ is a valid
$\frac{1}{3}$-deterministic QIVPM.

\amr{Explain significance}
